---
title: "CompareDAGs"
author: "Chloe Larkin"
date: "12/9/2020"
output: html_document
---

# Load bnlearn and graph visualization libraries
```{r}
library(bnlearn);
library(Rgraphviz);
```

# Read in the COVID data
```{r}
d_sep <- bnlearn:::dseparation
bnlearn_data = read.delim("../datasets/dag_data_V3.csv", sep = ",");
```

# Construct eight different experimental DAGs, which we will test Faithfulness and Markov assumptions on
```{r}
dag_1 = model2network("[CPL][HS][MI][SPL][UI][BLG|SPL][ICU|UI][CDC|HS:CPL:MI][CC|ICU:CDC:BLG]")
graphviz.plot(dag_1)
```

```{r}
dag_2 = model2network("[CPL|HS][HS][MI|HS][CDC|CPL:MI][CC|CDC:BLG][BLG|SPL][SPL]")
graphviz.plot(dag_2)
```

```{r}
dag_3 = model2network("[CC|CDC:BLG][CDC|HS][BLG|CPL:SPL][CPL|MI][SPL|MI][MI][HS]")
graphviz.plot(dag_3)
```


```{r}
dag_4 = model2network("[CC|CDC:CPL:BLG][BLG|SPL][SPL|CPL:MI][MI][CDC|CPL][CPL]")
graphviz.plot(dag_4)
```

```{r}
dag_5 = model2network("[CC|CDC:BLG][CDC|CPL][BLG|SPL][SPL|CPL][CPL|HS:MI][MI][HS]")
graphviz.plot(dag_5)
```

```{r}
dag_6 = model2network("[CC|CDC:BLG][CDC|CPL][BLG|SPL:UI][SPL|CPL][CPL|HS:MI][MI][HS][UI]")
graphviz.plot(dag_6)
```

```{r}
dag_7 = model2network("[CC|CDC:BLG][CDC|CPL][BLG|SPL][SPL|CPL][CPL|HS:UI][HS][UI]")
graphviz.plot(dag_7)
```
```{r}
dag_8 = model2network("[CC|CDC:BLG][CDC|CPL][BLG|SPL][SPL|CPL][CPL|HS:MI][HS][MI]")
graphviz.plot(dag_8)
```

# Method to get all sets of X, Y, and Z sets (including empty Z sets) for d-separation statements.
```{r}
get_args <- function(dag) {
    
    vars <- nodes(dag)
    pairs <- combn(x = vars, 2, list)
    true_dsep_statements <- list()
    nonredundant_dsep_statements <- list()
    arg_sets <- list()

    # Iterate through each pair of nodes in the DAG
    for(pair in pairs) {
        
        others <- setdiff(vars, pair)

        # Iterate through all possible Z sets for d-separating the current pair
        conditioning_sets <- unlist(lapply(0:4, function(.x) combn(others, .x, list)), recursive = F)
        for(set in conditioning_sets){
            xyz <- list(x = pair[1], y = pair[2], z = set)
            arg_sets <- c(arg_sets, list(xyz))

            # Check whether this X, Y, Z statement is truly d-separated
            if (d_sep(bn=dag, xyz[[1]], xyz[[2]], xyz[[3]])) {

                # If true d-separation statement, add to list of true d-separation statements
                true_dsep_statements <- c(true_dsep_statements, list(xyz));

                nonredundant = TRUE;

                # Check v-structure d-separation statements, where Z may be empty
                if(length(xyz[[3]]) == 0){ 
                    nonredundant = TRUE;
                }
                else{

                # Iterate over the node set Z: 
                    for(i in 1:length(xyz[[3]])){

                # Check d-separation of each Z subset, to see if the set is redundant:
                        if (d_sep(bn=dag, xyz[[1]], xyz[[2]], xyz[[3]][-i])){ 
                            nonredundant = FALSE;
                        }
                    }
                }

                # If not redundant, then add to list of nonredundant d-separation statements
                if(nonredundant){
                    nonredundant_dsep_statements <- c(nonredundant_dsep_statements, list(xyz));
                }
            }
        }
    }
    sprintf("Number of all true d-separation statements: %d; number of true and non-redundant d-separation statements: %d",
        length(true_dsep_statements), length(nonredundant_dsep_statements))
                                         
    return_vals <- list(arg_sets, true_dsep_statements)
                                 
    return(return_vals)
                                         
}
```

# Method to get the percentage of true d-separation statements that are also true conditional independence statements (Markov property)
```{r}
get_markov <- function(dag, true_dsep_statements) {
    true_cond_indi_count = 0;
    true_cond_indi_list <- list()

    # Iterate through each true d-separation statement
    for(statement in true_dsep_statements){

        # Check that the null hypothesis of conditional independence cannot be rejected 
        ci_test_results = ci.test(statement[[1]], statement[[2]], statement[[3]], data=bnlearn_data); 

        # Check the p-value of the null hypothesis rejection
        if(ci_test_results[2] >= 0.05){ 
            true_cond_indi_list <- c(true_cond_indi_list, ci_test_results$data.name)
            true_cond_indi_count = true_cond_indi_count + 1;
        }
    }
    if (length(true_dsep_statements) > 0) {
          return(true_cond_indi_count/length(true_dsep_statements)*100)
    }
    return (0)
}
```

# Method to get the percentage of true conditional independence statements that are also true d-separation statements (Faithfulness property)
```{r}
get_faithfulness <- function(dag, arg_sets) {
  # Create a list to store true conditional statements 
  true_conditional_statements <- list()
  true_cond_and_dsep <- list()
  
  # Evaluate whether each variable combination X, Y, {Z} in this DAG produces a true conditional statement
  for(set in arg_sets){
      # Evaluate the p-value of the ci.test result for this combination of X, Y, {Z}
      if (ci.test(set[[1]], set[[2]], set[[3]], bnlearn_data)[2] >= 0.05) { 
          # Add the set to the list of true conditional statements
          true_conditional_statements <- c(true_conditional_statements, list(set)) 
          
          # If the set is d-separated, dd the set to the list of true conditional and true d-separation statements
          if (d_sep(bn=dag, set[[1]], set[[2]], set[[3]])) { 
              true_cond_and_dsep = c(true_cond_and_dsep, list(set));
          }
      } 
  }
  if (length(true_conditional_statements) > 0) {
      return(length(true_cond_and_dsep)/length(true_conditional_statements) * 100)
  }
  return(0)
  
}
```

# Print each DAG's Markov and Faithfulness property performances
```{r}
dags <- list(dag_1, dag_2, dag_3, dag_4, dag_5, dag_6, dag_7, dag_8)

num = 1
for (dag in dags) {
  args = get_args(dag)
  markov_percent = get_markov(dag, args[[2]])
  faithfulness_percent = get_faithfulness(dag, args[[1]])
  print("DAG num, Markov %, Faithfulness %:")
  print(num)
  print(markov_percent)
  print(faithfulness_percent)
  num = num + 1
}


```

